EGGs are adhesive!

### 2025-03-03 calco/main.tex
miinima

\begin{abstract}
The use of rewriting-based visual formalisms is on the rise. 
%
In the formal methods community, this is due also to the introduction of adhesive
categories, where most properties of classical approaches to graph transformation, 
such as those on parallelism and causality, can be rephrased and proved in a general and 
uniform way.
%
E-graphs (or EGGs) are a formalism for program optimisation 
via an efficient implementation of equality saturation. 
In short, EGGs can be  defined as (acyclic) term graphs with an additional notion of 
equivalence on nodes that is closed under the operators of the signature.
Instead of replacing the components of a program, the optimisation step 
is performed by adding new components and linking them to 
the existing ones via an equivalence relation, until an optimal program is reached.
%
This work describes EGGs by adhesive categories. Besides 
the benefits \emph{per se} of a formal presentation, making precise the properties of 
the data structure, describing the addition of equivalent program components 
via the standard tools of graph transformation offers the advantages given by the 
adhesive framework in modelling e.g. concurrent updates.
\end{abstract}

@@

\emph{Synopsis}
The paper has the following structure. 
In Section~\ref{sec:ade} we briefly recall 
the theory of $\mathcal{M}$-adhesive categories
and of kernel pairs.
In Section~\ref{sec:hyper} we present the graphical structures of our interest, 
 (labelled, acyclic) hypergraphs and term graphs, and we provide a
functorial characterisation, which allows for proving their adhesivity properties.
This is expanded in Section~\ref{hypereq} for describing hypergraphs 
and term graphs with equivalence and in Section~\ref{eggs} for capturing
their variants where the equivalences are closed with respect to operator application,
thus subsuming EGGs.
%
Finally, in Section~\ref{conclusioni} we show draw our conclusions: we poi to further work
and we show how optimisation steps can ve described via DPO transformation.
%
For the sake of space, all the proofs have been moved to the appendices. 
%For the sake of completeness, 
%and in order to fix the notation, we prove all the results 
%recalled in the background section, besides those that are original 
%to our work.

@@

\section{Pros and cons of adhesive rewriting}
The previous sections has shown how term graphs with (possibly) closed equivalence
can be described as a suitable $\mathcal{M},\mathcal{N}$-adhesive category. 
The latter precisely capture EGGS, as originally
presented in~\cite{WillseyNWFTP21}. 
%
Of course, being a purely graphical representation, there can be redundancy and possibly ambiguity,
so that e.g. a term such as $f(a,a)$ admits a few different representations as a term graphs, given below

>>>>

\begin{abstract}
The use of rewriting-based visual formalisms is on the rise. 
%
In the formal methods community, this is due also to the introduction of adhesive
categories, where most properties of classical approaches to graph transformation, 
such as those on parallelism and causality, can be rephrased and proved in a general and 
uniform way.
%
E-graphs (EGGs) are a formalism for program optimisation 
via an efficient implementation of equality saturation. 
In short, EGGs can be  defined as (acyclic) term graphs with an additional notion of 
equivalence on nodes that is closed under the operators of the signature.
Instead of replacing the components of a program, the optimisation step 
is performed by adding new components and linking them to 
the existing ones via an equivalence relation, until an optimal program is reached.
%
This work describes EGGs via adhesive categories. 
Besides the benefits in itself of a formal presentation, which renders the 
properties of the data structure precise, the description of the addition of equivalent 
program components using standard graph transformation tools offers the advantages 
of the adhesive framework in modelling, for example, concurrent updates.
%
%Besides 
%the benefits \emph{per se} of a formal presentation, making precise the properties of 
%the data structure, describing the addition of equivalent program components 
%via the standard tools of graph transformation offers the advantages given by the 
%adhesive framework in modelling e.g. concurrent updates.
\end{abstract}

@@

\emph{Synopsis}
The paper has the following structure. 
In Section~\ref{sec:ade} we briefly recall 
the theory of $\mathcal{M}$-adhesive categories
and of kernel pairs.
In Section~\ref{sec:hyper} we present the graphical structures of our interest, 
 (labelled, acyclic) hypergraphs and term graphs, and we provide a
functorial characterisation, which allows for proving their adhesivity properties.
This is expanded in Section~\ref{hypereq} for describing hypergraphs 
and term graphs with equivalence and in Section~\ref{eggs} for capturing
their variants where the equivalences are closed with respect to operator application,
thus subsuming EGGs.
%
In Section~\ref{rewriting} we put the former machinery at work, showing how the optimisation steps
can be rephrased as suitable applications of term graph rewriting rules.
%
Finally, in Section~\ref{conclusioni} we draw our conclusions, hint at future endeavours and offer some 
brief remarks on related works.
%
For the sake of space, proofs have been moved to the appendices. 
%For the sake of completeness, 
%and in order to fix the notation, we prove all the results 
%recalled in the background section, besides those that are original 
%to our work.

@@

\section{Pros and cons of adhesive rewriting}
\label{rewriting}
The previous sections has shown how term graphs with (possibly) closed equivalence
can be described as a suitable $\mathcal{M},\mathcal{N}$-adhesive category. 
The latter precisely capture EGGS, as originally
presented in~\cite{WillseyNWFTP21}. 
%
Of course, being a purely graphical representation, there can be redundancy and possibly ambiguity,
so that e.g. a term such as $f(a,a)$ admits a few different representations as a term graphs, given below
